% dedication text

Thaumaturgike, is that Art Mathematicall, which giueth certaine order to make straunge workes, of the sense to be perceiued, and of men greatly to be wondred at. By sundry meanes, this Wonder-worke is wrought. Some, by Pneumatithmie. As the workes of Ctesibius and Hero, some by waight wherof Timæus speaketh. Some, by Stringes strayned, or Springs, therwith Imitating liuely Motions. Some, by other meanes, as the Images of Mercurie: and the brasen hed, made by Albertus Magnus, which dyd seme to speake. Boethius was excellent in these feates. To whom, Cassiodorus writyng, sayth. Your purpose is to know profound thynges: and to shew meruayles. By the disposition of your Arte, Metals do low: Diomedes of brasse, doth blow a Trumpet loude: a brasen Serpent hisseth: byrdes made, sing swetely. Small thynges we rehearse of you, who can Imitate the heauen.

\phantom{text} \hfill — John Dee, 1570

(The Mathematicall Praeface to the Elements of Geometrie of Euclid of Megara)

\vspace{2cm}

``It always seemed to me they’re sort of alike,'' he said, ``magic and music.
Spells and tunes. For one thing, you have to get them just exactly right.''

\phantom{text} \hfill — Ursula K. Le Guin, 2001

\phantom{text} \hfill (Tales from Earthsea)

\vspace{2cm}

[We] creators, are we not, in the end, people of the greatest \emph{faith}? The greatest mystics of our time. If I say yes, I fully realize that that does not take us forward a single step. [...] Creation, in its æsthetic necessity, remains incomprehensible, for one knows very well that it's not enough to make rows of \lilyDynamics{f} and \lilyDynamics{p} markings in order to make a worthwhile work, but from the moment ``something sparks'' one enters into a domain as senseless as that of a rock becoming man, where a man leaves behind rationality involuntarily to enter into the irrational, becoming mad. Almost nobody has \emph{felt} the troubling temperature of our time, our historical truth. They do not know that, more than ever, the only \emph{possible} thing is a total seizing of the universe, and that without that anything is not only incomplete but \emph{useless}.

\phantom{text} \hfill — Jean Barraqué, 1950's

\phantom{text} \hfill (Personal Correspondence)