% the acknowledgments section
\doublespacing

\newthought{\hspace{1cm} I would like to acknowledge} the support of my advisor David Gompper. Thank you for all the re-directions of my research and creative work: always exactly when I needed it. Thank you for having a wonderful ear for harmony and form and for supporting many of the experimental paths I've chosen over the last few years. ❧ Thanks to Sarah Suhadolnik for the many sessions tightening my prose and helping clarify my points. Your enthusiasm and confidence in my project carried me through many doubts and hesitations. I left every meeting energized and excited to carry on. ❧ Thank you Joshua Albrecht for joining the committee of such a surprisingly large project at the last-minute. ❧ Trevor Bača for encouraging my passion for formalization and training me in computation and Rourke O'Brien for letting me borrow him so often. Through you I have perhaps learned the most about myself (and a thing or two about semicolons). Here's to many more espresso-fueled adventures. ❧ Trinton Hlynn for all the chats and consultations and for filling the air during so many drives and hikes. ❧ Joséphine Wolf Oberholtzer and Jeff Treviño for all the help with Abjad in the early days. I'm not sorry for filling your inboxes. ❧ Chris Otto for leading me to non-tempered tunings. ❧ Liam Battle for continued excitment about all of my projects. ❧ José Besada for pointing me to important resources. ❧ My parents Dale and Sandra for their constant support. I would have not come so far if it wasn't for you fostering my interests after falling in love with music at the Dayton Philharmonic. ❧ My brother Geoff for every diversion, road tripping to see movies, and listening to music together. ❧ and Stevie Haley Delgado for the debates, sculptures, shadows, and bunnies. And also for Salem.

% \cleardoublepage